Joint maximum-likelihood of phylogenies and ancestral states is not consistent

### 2018-05-07 appendix.tex
unique maximum! and a little rearranging

We state the Hadamard representation of site split generating probabilities---that is, probabilities of obtaining particular site splits given a tree---following Section 8.6 of \citet{Semple2003-em}.
For each edge $e$ define the edge ``fidelity'' for that edge as
\[
\theta(e) = 1-2p(e)
\]
where $p(e)$ is the probability of a character change along edge $e$.
For an even-sized subset of $Y\subseteq\mathcal{S}$, let the path set $P(Y)$ be the set of edges in the path connecting both elements of $Y$.
For $n$ taxa, the probability of observing site split $A\in\mathcal{P}(\siteSplitSet)$ is
\begin{equation}
\label{eq:hadamard_probability}
p_A = \frac{1}{2^{n-1}} \ \sum_{Y \subseteq \mathcal{S} : |Y| \equiv 0 (\mathrm{mod} \ 2)} \ \left[(-1)^{|Y \cap A|} \ \prod_{e\in P(Y)} \ \theta(e) \right].
\end{equation}
By convention, we set $P(\emptyset)=\emptyset$ and $\prod_{e\in\emptyset} \ \theta(e) = 1$.
For notational convenience, let
\[
p_{\siteSplit_j} := \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,t),
\]
for any site split $\siteSplit_j$.
Table~\ref{tab:sitepatprob} contains calculations of site pattern probabilities for our two topologies (Fig.~\ref{fig:farris-fels-top}).

\begin{table}[ht]
\centering
\begin{tabular}{|l|l|l|l|}
    \hline
$\siteSplit_j$  & $p_{\siteSplit_j}$ &$\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,t)$&$\Pr(\siteSplitRV=\siteSplit_j \mid \tau_2,t)$\\
    \hline
    $\emptyset$ & $p_{\emptyset}$   &$1+x^2+y^2+4xyw+x^2y^2$&$1+2xy+2xyw+x^2w+y^2w+x^2y^2$\\
    $\{1\}$     & $p_{1}$   &$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
    $\{2\}$     & $p_{2}$   &$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
    $\{3\}$     & $p_{3}$   &$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
    $\{1,2\}$   & $p_{12}$   &$1-x^2-y^2+x^2y^2$&$1+2xy-2xyw-x^2w-y^2w+x^2y^2$\\
    $\{1,3\}$   & $p_{13}$   &$1+x^2+y^2-4xyw+x^2y^2$&$1-2xy-2xyw+x^2w+y^2w+x^2y^2$\\
    $\{2,3\}$   & $p_{23}$   &$1-x^2-y^2+x^2y^2$&$1-2xy+2xyw-x^2w-y^2w+x^2y^2$\\
    $\{1,2,3\}$ & $p_{123}$   &$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
    \hline
\end{tabular}
\caption{Site pattern probabilities $p_{\siteSplit_j}$ on the InvFels tree $\tau_1$ and the Felsenstein tree $\tau_2$ obtained using the Hadamard transform.
All values multiplied by $1/8$.}
\label{tab:sitepatprob}
\end{table}

@@

Since exchanging $x_1$ and $x_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t})$, if there is a unique maximum of the log-likelihood we will have $x_1=x_2$ at the maximum.
An analogous statement holds for $y_1$ and $y_2$.
It remains to show that the joint inference procedure maximizing \eqref{eq:profile_likelihood} results in a unique estimate of $t$.
Without loss of generality, we focus on the single parameter $x_1$ and show we obtain a unique maximum when performing joint inference; similar arguments hold for the remaining parameters.
We decompose the likelihood into the entropy and partial log-likelihood terms as in \eqref{eq:log_likelihood_simplified}.
%EM I'm thinking about this wrong for some reason: the sum of two things, each of which has a unique maximum, does not necessarily have a unique maximum.
By Gibbs's inequality, the entropy term $\shannonDivergence_{\tau^*,t^*}(\tau^*,t)$ has a unique maximum over $t$---namely $t^*$.
%EM There's not good notational overlap between Table S2 and the definition of partial log-likelihood. Could you fix that?
%das: I think fixed, or did you mean something in addition to changing h-->\Xi and y-->\Psi?
For the general case of $\tilde{t}=\{x_1, y_2, x_1, y_2, w\}$, we can compute the partial log-likelihood $\tilde{\ell}_{\tau^*,t^*}(\tau^*, t)$ as in Table~\ref{tab:farris_likelihoods}.
Clearly this partial log-likelihood will be of the form $g(x_1) = p\log(1+x_1) + (1-p)\log(1-x_1)$ with $0 \le p \le 1$, which, by the second derivative test, is concave for $x_1\in[0,1]$ and thus has a unique maximum.
Since both summands of the likelihood are concave on $x_1\in[0,1]$, the likelihood is as well, and there exists a single $x_1\in[0,1]$ maximizing the likelihood.
Given this and the symmetry of the likelihood, we let $x_1=x_2=x$ and $y_1=y_2=y$.
The Felsenstein tree does not admit this property, but, since we are interested in a lower bound for this tree, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.

@@

\begin{proof}
In general,
\[
\max_{t} \ \ell_{\tau^*,t^*}(\tau, t) \le
    \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
    + \tilde{\ell}_{\tau^*,t^*}(\tau, \hat{t})
\]
is an upper bound for the joint maximum of \eqref{eq:log_likelihood_simplified} using Gibbs's inequality
\[
\shannonDivergence_{\tau^*,t^*}(\tau,t) \le \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
\]
and
\[
\hat{t} = \argmax_{t} \ \tilde{\ell}_{\tau^*,t^*}(\tau, t).
\]
Similarly,
\[
\max_{t} \ \ell_{\tau^*,t^*}(\tau, t) \ge
    \shannonDivergence_{\tau^*,t^*}(\tau,\hat{t})
    + \tilde{\ell}_{\tau^*,t^*}(\tau, \hat{t})
\]
is a lower bound.

Assume $\tau^*=\tau_1$.
The InvFels tree log-likelihood takes one of three values depending on branch lengths, which is due to site split $\{1,3\}$ (see the upper table of Table~\ref{tab:likelihoods}).

@@

We next construct a lower bound on the Felsenstein tree partial likelihood.
In the Felsenstein case, there are many more optimal internal states depending on branch lengths than in the InvFels case (see the lower table of Table~\ref{tab:likelihoods}).
We first bound the entire likelihood below and then proceed with joint inference.
To obtain a lower bound for the likelihood we replace $1+w$ with $1-w$ in Table~\ref{tab:likelihoods} for $\tau = \tau_2$; in this case, we resolve many of the ambiguous likelihood terms.
For example, for site split $\{1,2\}$, after we replace the $1+w$ term for $\ancestralSplitPartition_j(\tau, t)=\{\emptyset, \{1,2\}\}$ with $1-w$, we have
\[
\max\left((1+x)^2 (1-w)(1+y)^2, (1+x)(1-x)(1-w)(1+y)(1-y)\right)
\]
where since
\[
(1+x)^2 (1-w)(1+y)^2 > (1+x)(1-x)(1-w)(1+y)(1-y),
\]
the maximum for this site split can be bounded below by $(1+x)^2 (1-w)(1+y)^2$.
For the other site splits we consider two cases.
If $(1+x)(1-y) > (1-x)(1+y)$, then
\[
(1+x)^2(1-w)(1+y)(1-y) > (1+x)(1-x)(1-w)(1+y)^2
\]
meaning the cases $\{1\}, \{2\}, \{3\}$ and $\{1,2,3\}$ will have likelihoods bounded below by
\[
(1+x)^2(1-w)(1+y)(1-y).
\]
For the cases $\{1,3\}$ and $\{2,3\}$ and the same condition, we have
\[
(1+x)^2(1-w)(1-y)^2 > (1-x)^2(1-w)(1+y)^2
\]
and
\[
(1+x)^2(1-w)(1-y)^2 > (1+x)(1-x)(1-w)(1+y)(1-y),
\]
yielding lower bounds of
\[
(1+x)^2(1-w)(1-y)^2
\]
in these cases.
The partial likelihood in this case is then bounded below by
\begin{align*}
    \tilde{\ell}_{\tau_1,t^*}(\tau_2, t)
    &\ge      p_{\emptyset}  \cdot\log((1+x)^2   (1-w)(1+y)^2) \\
    &\qquad + p_{1}          \cdot\log((1+x)^2   (1-w)(1+y)(1-y)) \\
    &\qquad + p_{2}          \cdot\log((1+x)^2   (1-w)(1+y)(1-y)) \\
    &\qquad + p_{3}          \cdot\log((1+x)^2   (1-w)(1+y)(1-y)) \\
    &\qquad + p_{12}         \cdot\log((1+x)^2   (1-w)(1+y)^2) \\
    &\qquad + p_{123}        \cdot\log((1+x)^2   (1-w)(1+y)(1-y))\\
    &\qquad + p_{13}         \cdot\log((1+x)^2   (1-w)(1-y)^2) \\
    &\qquad + p_{23}         \cdot\log((1+x)^2   (1-w)(1-y)^2).
\end{align*}
When $(1+x)(1-y) < (1-x)(1+y)$, similar arguments yield the lower bound
\begin{align*}
    \tilde{\ell}_{\tau_1,t^*}(\tau_2, t)
    &\ge      p_{\emptyset}  \cdot\log((1+x)^2    (1-w)(1+y)^2) \\
    &\qquad + p_{1}          \cdot\log((1+x)(1-x)(1-w)(1+y)^2) \\
    &\qquad + p_{2}          \cdot\log((1+x)(1-x)(1-w)(1+y)^2) \\
    &\qquad + p_{3}          \cdot\log((1+x)(1-x)(1-w)(1+y)^2) \\
    %See above.
    &\qquad + p_{12}         \cdot\log((1+x)^2    (1-w)(1+y)^2) \\
    &\qquad + p_{123}        \cdot\log((1+x)(1-x)(1-w)(1+y)^2)\\
    &\qquad + p_{13}         \cdot\log((1-x)^2    (1-w)(1+y)^2) \\
    &\qquad + p_{23}         \cdot\log((1-x)^2    (1-w)(1+y)^2).
\end{align*}
Letting
\[
b = p_{1}+p_{2}+p_{3}+p_{123}
\]
yields
\[
    \tilde{\ell}_{\tau_1,t^*}(\tau_2, t)
    \ge      2\cdot\log(1+x)
    + (2-b)  \cdot\log(1+y)
    + b      \cdot\log(1-y)
    + \log(1-w)
\]
for the first lower bound, and we get the equivalent expression for the second lower bound by exchanging $x$ and $y$.
Maximizing either lower bound over $t$, we obtain
\[
\tilde{\ell}_{\tau_1,t^*}(\tau_2, t)
\ge      2\cdot\log(2)
+ (2-b)  \cdot\log(2-b)
+ b      \cdot\log(b),
\]
i.e., the maximum is achieved at $\{1,1-b,0\}$.
Our lower bound is
\[
C_1(x^*, y^*) := \shannonDivergence_{\tau_1,t^*}(\tau_2,\{1,1-b,0\}) + 2\cdot\log(2) + (2-b)\cdot\log(2-b) + b\cdot\log(b).
\]
See Fig.~\ref{fig:inconsistency-farris} for the values of $0 < x^*, y^* < 1$ where $C_1(x^*, y^*) > C_0(x^*, y^*)$ and joint inference is inconsistent.
\end{proof}

>>>>

We state the Hadamard representation of site split generating probabilities---that is, probabilities of obtaining particular site splits given a tree---following Section 8.6 of \citet{Semple2003-em}.
For each edge $e$ define the edge ``fidelity'' for that edge as
\[
\theta(e) = 1-2p(e)
\]
where $p(e)$ is the probability of a character change along edge $e$.
For an even-sized subset of $Y\subseteq\mathcal{S}$, let the path set $P(Y)$ be the set of edges in the path connecting both elements of $Y$.
For $n$ taxa, the probability of observing site split $A\in\mathcal{P}(\siteSplitSet)$ is
\begin{equation}
\label{eq:hadamard_probability}
p_A = \frac{1}{2^{n-1}} \ \sum_{Y \subseteq \mathcal{S} : |Y| \equiv 0 (\mathrm{mod} \ 2)} \ \left[(-1)^{|Y \cap A|} \ \prod_{e\in P(Y)} \ \theta(e) \right].
\end{equation}
By convention, we set $P(\emptyset)=\emptyset$ and $\prod_{e\in\emptyset} \ \theta(e) = 1$.
For notational convenience, let
\[
p_{\siteSplit_j} := \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,t),
\]
for any site split $\siteSplit_j$.
Table~\ref{tab:gen-sitepatprob} contains calculations of site pattern probabilities for the InvFels topology with arbitrary branch lengths and Table~\ref{tab:sitepatprob} site pattern probabilities for the InvFels topology in Fig.~\ref{fig:farris-fels-top}a.

\begin{table}[ht]
\centering
\begin{tabular}{|l|l|l|}
    \hline
$\siteSplit_j$  & $p_{\siteSplit_j}$ &$\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,\{x_1,y_1,x_2,y_2,w\})$\\
    \hline
    $\emptyset$ & $p_{\emptyset}$   &$1 + x_1x_2 +  y_1y_2 +  w[  x_1 + x_2][  y_1 + y_2] + x_1y_1x_2y_2$\\
    $\{1\}$     & $p_{\{1\}}$       &$1 - x_1x_2 +  y_1y_2 +  w[- x_1 + x_2][  y_1 + y_2] - x_1y_1x_2y_2$\\
    $\{2\}$     & $p_{\{2\}}$       &$1 + x_1x_2 -  y_1y_2 +  w[  x_1 + x_2][- y_1 + y_2] - x_1y_1x_2y_2$\\
    $\{3\}$     & $p_{\{3\}}$       &$1 - x_1x_2 +  y_1y_2 +  w[  x_1 - x_2][  y_1 + y_2] - x_1y_1x_2y_2$\\
    $\{1,2\}$   & $p_{\{1,2\}}$     &$1 - x_1x_2 -  y_1y_2 +  w[- x_1 + x_2][- y_1 + y_2] + x_1y_1x_2y_2$\\
    $\{1,3\}$   & $p_{\{1,3\}}$     &$1 + x_1x_2 +  y_1y_2 +  w[- x_1 - x_2][  y_1 + y_2] + x_1y_1x_2y_2$\\
    $\{2,3\}$   & $p_{\{2,3\}}$     &$1 - x_1x_2 -  y_1y_2 +  w[  x_1 - x_2][- y_1 + y_2] + x_1y_1x_2y_2$\\
    $\{1,2,3\}$ & $p_{\{1,2,3\}}$   &$1 + x_1x_2 -  y_1y_2 +  w[  x_1 + x_2][  y_1 - y_2] - x_1y_1x_2y_2$\\
    \hline
\end{tabular}
\caption{Site pattern probabilities $p_{\siteSplit_j}$ on the InvFels tree $\tau_1$ with $t=\{x_1,y_1,x_2,y_2,w\}$ obtained using the Hadamard transform.
All values multiplied by $1/8$.}
\label{tab:gen-sitepatprob}
\end{table}

\begin{table}[ht]
\centering
\begin{tabular}{|l|l|l|}
    \hline
$\siteSplit_j$  & $p_{\siteSplit_j}$ &$\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,\{x,y,x,y,w\})$\\
    \hline
    $\emptyset$ & $p_{\emptyset}$   &$1+x^2+y^2+4xyw+x^2y^2$\\
    $\{1\}$     & $p_{1}$   &$1-x^2+y^2-x^2y^2$\\
    $\{2\}$     & $p_{2}$   &$1+x^2-y^2-x^2y^2$\\
    $\{3\}$     & $p_{3}$   &$1-x^2+y^2-x^2y^2$\\
    $\{1,2\}$   & $p_{12}$   &$1-x^2-y^2+x^2y^2$\\
    $\{1,3\}$   & $p_{13}$   &$1+x^2+y^2-4xyw+x^2y^2$\\
    $\{2,3\}$   & $p_{23}$   &$1-x^2-y^2+x^2y^2$\\
    $\{1,2,3\}$ & $p_{123}$   &$1+x^2-y^2-x^2y^2$\\
    \hline
\end{tabular}
\caption{Site pattern probabilities $p_{\siteSplit_j}$ on the InvFels tree $\tau_1$ with $t=\{x,y,x,y,w\}$ obtained using the Hadamard transform.
All values multiplied by $1/8$.}
\label{tab:sitepatprob}
\end{table}

@@

Since exchanging $x_1$ and $x_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t})$, if there is a unique maximum of the log-likelihood we will have $x_1=x_2$ at the maximum.
An analogous statement holds for $y_1$ and $y_2$.
It remains to show that the joint inference procedure maximizing \eqref{eq:profile_likelihood} results in a unique estimate of $t$.
Without loss of generality, we focus on the single parameter $x_1$ and show we obtain a unique maximum when performing joint inference; similar arguments hold for the remaining parameters.
We decompose the likelihood into the entropy and partial log-likelihood terms as in \eqref{eq:log_likelihood_simplified}.
%EM I'm thinking about this wrong for some reason: the sum of two things, each of which has a unique maximum, does not necessarily have a unique maximum.
%das: I added the table of general generating probabilities, but it's pretty straightforward I think to see below the second derivative of the likelihood will be negative on the whole interval, implying we have a unique maximum.
%I think we're good.
%Let me know if I should say anything about the continuity of the likelihood, but I think it's obvious this gets us to where we need to be?
Consider the general case of $\tilde{t}=\{x_1, y_2, x_1, y_2, w\}$, and the map $g(x_1) \mapsto -g(x_1)\log g(x_1)$.
We restrict our attention to $x_1,y_1,x_2,y_2,w\in(0,1)$, i.e., cases where $0<p_{\siteSplit_j}<1$ for all $\siteSplit_j$.
By straightforward calculus, the second derivative of this map is
\[
-g''(x_1)(1+\log g(x_1)) - \frac{g'(x_1)^2}{g(x_1)}.
\]
Using each functional form in Table~\ref{tab:gen-sitepatprob} we see in all cases $g''(x_1)=0$ and $g(x_1) > 0$ for $x_1\in(0,1).$
This implies the above second derivative is negative for each component of $\shannonDivergence_{\tau^*,t^*}(\tau^*,\tilde{t})$, and the second derivative of $\shannonDivergence_{\tau^*,t^*}(\tau^*,\tilde{t})$ itself is negative.
%EM There's not good notational overlap between Table S2 and the definition of partial log-likelihood. Could you fix that?
%das: I think fixed, or did you mean something in addition to changing h-->\Xi and y-->\Psi?

We can compute the partial log-likelihood $\tilde{\ell}_{\tau^*,t^*}(\tau^*, t)$ as in Table~\ref{tab:farris_likelihoods}.
Clearly this partial log-likelihood will be of the form $g(x_1) = p\log(1+x_1) + (1-p)\log(1-x_1)$ with $0 \le p \le 1$, which, by the second derivative test, is concave for $x_1\in(0,1)$ and thus has a unique maximum.
Since both summands of the likelihood are concave on $x_1\in(0,1)$, the likelihood is as well, and there exists a single $x_1\in(0,1)$ maximizing the likelihood.
Given this and the symmetry of the likelihood, we let $x_1=x_2=x$ and $y_1=y_2=y$.
The Felsenstein tree does not admit this property, but, since we are interested in a lower bound for this tree, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.

@@

\begin{proof}
Assume $\tau^*=\tau_1$.
The InvFels tree log-likelihood takes one of three values depending on branch lengths, which is due to site split $\{1,3\}$ (see the upper table of Table~\ref{tab:likelihoods}).

@@

We next construct a lower bound on the Felsenstein tree partial likelihood.
%das: we never actually use the Gibbs inequality trick for the upper bound---just some log-of-polynomial bounds
In general, taking
\hat{t} = \argmax_{t} \ \tilde{\ell}_{\tau^*,t^*}(\tau, t).
we see
\max_{t} \ \ell_{\tau^*,t^*}(\tau, t) \ge
    \shannonDivergence_{\tau^*,t^*}(\tau,\hat{t})
    + \tilde{\ell}_{\tau^*,t^*}(\tau, \hat{t})
is a lower bound for the joint maximum of \eqref{eq:log_likelihood_simplified}.
In the Felsenstein case, there are many more optimal internal states depending on branch lengths than in the InvFels case (see the lower table of Table~\ref{tab:likelihoods}).
We first bound the entire likelihood below and then proceed with joint inference.
To obtain a lower bound for the likelihood we replace $1+w$ with $1-w$ in Table~\ref{tab:likelihoods} for $\tau = \tau_2$; in this case, we resolve many of the ambiguous likelihood terms.
For example, for site split $\{1,2\}$, after we replace the $1+w$ term for $\ancestralSplitPartition_j(\tau, t)=\{\emptyset, \{1,2\}\}$ with $1-w$, we have
\max\left((1+x)^2 (1-w)(1+y)^2, (1+x)(1-x)(1-w)(1+y)(1-y)\right)
where since
(1+x)^2 (1-w)(1+y)^2 > (1+x)(1-x)(1-w)(1+y)(1-y),
the maximum for this site split can be bounded below by $(1+x)^2 (1-w)(1+y)^2$.
For the other site splits we consider two cases.
If $(1+x)(1-y) > (1-x)(1+y)$, then
(1+x)^2(1-w)(1+y)(1-y) > (1+x)(1-x)(1-w)(1+y)^2
meaning the cases $\{1\}, \{2\}, \{3\}$ and $\{1,2,3\}$ will have likelihoods bounded below by
(1+x)^2(1-w)(1+y)(1-y).
For the cases $\{1,3\}$ and $\{2,3\}$ and the same condition, we have
(1+x)^2(1-w)(1-y)^2 > (1-x)^2(1-w)(1+y)^2
and
(1+x)^2(1-w)(1-y)^2 > (1+x)(1-x)(1-w)(1+y)(1-y),
yielding lower bounds of
(1+x)^2(1-w)(1-y)^2
in these cases.
The partial likelihood in this case is then bounded below by
    \tilde{\ell}_{\tau_1,t^*}(\tau_2, t)
    &\ge      p_{\emptyset}  \cdot\log((1+x)^2   (1-w)(1+y)^2) \\
    &\qquad + p_{1}          \cdot\log((1+x)^2   (1-w)(1+y)(1-y)) \\
    &\qquad + p_{2}          \cdot\log((1+x)^2   (1-w)(1+y)(1-y)) \\
    &\qquad + p_{3}          \cdot\log((1+x)^2   (1-w)(1+y)(1-y)) \\
    &\qquad + p_{12}         \cdot\log((1+x)^2   (1-w)(1+y)^2) \\
    &\qquad + p_{123}        \cdot\log((1+x)^2   (1-w)(1+y)(1-y))\\
    &\qquad + p_{13}         \cdot\log((1+x)^2   (1-w)(1-y)^2) \\
    &\qquad + p_{23}         \cdot\log((1+x)^2   (1-w)(1-y)^2).
When $(1+x)(1-y) < (1-x)(1+y)$, similar arguments yield the lower bound
    \tilde{\ell}_{\tau_1,t^*}(\tau_2, t)
    &\ge      p_{\emptyset}  \cdot\log((1+x)^2    (1-w)(1+y)^2) \\
    &\qquad + p_{1}          \cdot\log((1+x)(1-x)(1-w)(1+y)^2) \\
    &\qquad + p_{2}          \cdot\log((1+x)(1-x)(1-w)(1+y)^2) \\
    &\qquad + p_{3}          \cdot\log((1+x)(1-x)(1-w)(1+y)^2) \\
    %See above.
    &\qquad + p_{12}         \cdot\log((1+x)^2    (1-w)(1+y)^2) \\
    &\qquad + p_{123}        \cdot\log((1+x)(1-x)(1-w)(1+y)^2)\\
    &\qquad + p_{13}         \cdot\log((1-x)^2    (1-w)(1+y)^2) \\
    &\qquad + p_{23}         \cdot\log((1-x)^2    (1-w)(1+y)^2).
Letting
b = p_{1}+p_{2}+p_{3}+p_{123}
yields
    \tilde{\ell}_{\tau_1,t^*}(\tau_2, t)
    \ge      2\cdot\log(1+x)
    + (2-b)  \cdot\log(1+y)
    + b      \cdot\log(1-y)
    + \log(1-w)
for the first lower bound, and we get the equivalent expression for the second lower bound by exchanging $x$ and $y$.
Maximizing either lower bound over $t$, we obtain
\tilde{\ell}_{\tau_1,t^*}(\tau_2, t)
\ge      2\cdot\log(2)
+ (2-b)  \cdot\log(2-b)
+ b      \cdot\log(b),
i.e., the maximum is achieved at $\{1,1-b,0\}$.
Our lower bound is
C_1(x^*, y^*) := \shannonDivergence_{\tau_1,t^*}(\tau_2,\{1,1-b,0\}) + 2\cdot\log(2) + (2-b)\cdot\log(2-b) + b\cdot\log(b).
See Fig.~\ref{fig:inconsistency-farris} for the values of $0 < x^*, y^* < 1$ where $C_1(x^*, y^*) > C_0(x^*, y^*)$ and joint inference is inconsistent.
\end{proof}
